Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-10 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex

\begin{referee}
Concerning the source/sink term, I think that the usage of such an artificial term is relevant if
the effect of the term is small. Can you show the effect of the source/sink term is small or not in
some cases shown in Section VI, by comparing $\left< \int d^3v\, S(\psi,v) \right> \Delta t$ and $n(\psi)$, for example, where
$\Delta t$ is some characteristic time scale such as collision time or transport time scale.
\end{referee}

Response goes here.

\begin{referee}
Explanation for Fig 1 and 2: Though authors write ``As $E_r \to 0$, all the matrix elements converge
smoothly to their $E_r = 0$ limits'', since the horizontal axis of these figures are in log scale, we
cannot judge if their claim is true or not. Please show the value of $L_{jk}(E_r=0)$ in these figures,
too.
\end{referee}

The values for $E_r=0$ are now plotted on figures (1)-(2), as $\blacktriangleright$ symbols on the left axis
of each sub-figure.

@@

\begin{referee}
It is estimated that $E_* \sim 0.1 - 0.01$ in most of the plasma. However, if the core $T_e$ is higher than
$T_i$ and ambipolar $E_r$ is positive (electron root) in ECH-heated plasma, $E_*$ for ions easily becomes
nearly, or even larger than, unity near the magnetic axis, since $\rho_\theta \propto 1/B_p$ is large and usually
$|E_r|$(ion root) $\ll E_r$(electron root). See Yokoyama et al., Nuclear Fusion (2007), 1213 for example.
I think therefore the statement ``In most of the plasma, $E_*$ is however expected to be smaller
than a few percent.'' is too strong and may be misleading.
\end{referee}

Response goes here.

\begin{referee}
Appendix: The authors claims that the parallel flow and heat flux solved by SFINCS code obeys the
isomorphism, but is radial electric field ($d\Phi/dr$) contained in ``other input quantities''? Please
clarity that. Also, if the isomorphism is satisfied for finite $d\Phi/dr$ and finite source term, I
suppose that the usage of source term in the full trajectory model is justified at least in
quasisymmetric systems. Can author comment on this point in Appendix?
\end{referee}

We have changed and expanded the text in this appendix for clarification. 
For all of the trajectory models described, an isomorphism does hold for the transport matrix elements
if the electric field is varied so
$(GM+IN)(\iota M-N)^{-1} d\Phi_0/d\psi$ remains constant as $M$ and $N$ are varied.
%This required transformation of $d\Phi_0/d\psi$ can be found in the last line of table 1 in Ref. [49].
\vspace{0.3in}

### 2014-02-10 doc/sfincsPaper/sfincsPaper.tex
Added features to matlab 1-species version for testing quasisymmetry isomorphism and for non-stellarator-symmetric geometryScheme=1. Also expanded Appendix A of paper.

\begin{figure}[h!]
% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
%******************************
% Previous version of figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% m20130324_01_SFINCS_ErScan_2013-03-27_03-59_ErScan_awesomer_mergeder.mat
% Code run on Matt's laptop
% *********************************
\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_LHD.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.pdf}
\caption{(Color online) Comparison of trajectory models for LHD standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
of the normalized radial electric field (\ref{eq:EStar}).
\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
\label{fig:ErComparison_LHD}}
\end{figure}

\begin{figure}[h!]
% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
%******************************
% Previous version of figure generated using m20131205_05_plotSFINCSFortranErComparisonForPaper.m
% SFINCS run on hopper in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
%******************************
% Old version of figure from Dec 2 version:
% Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% sfincs_2013-12-01_15-49_ErScan_noConvergence
% Code run on edison in 
% *********************************
\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_W7X.eps}
%\includegraphics{m20131205_05_plotSFINCSFortranErComparisonForPaper_W7X.eps}
\caption{(Color online) Comparison of trajectory models for W7-X standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
of the normalized radial electric field (\ref{eq:EStar}).
\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
\label{fig:ErComparison_W7X}}
\end{figure}

@@

A useful test of a stellarator neoclassical code such as the one
described here is the quasisymmetry isomorphism,
discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
In this test, the magnetic field is specified as $B(\theta,\zeta) = y(M\theta-N\zeta)$
for some given periodic function $y$ and integers $M$ and $N$.
The kinetic equation is solved for several values of $M$ and $N$,
varying the collision frequency with each run so that
$\nuii / \left|M - N/\iota \right|$ remains fixed.
As $M$ and $N$ are varied, $G$, $I$, $\iota$, $dn/d\psi$, $dT/d\psi$, and other input quantities are also kept fixed.
Under these circumstances, it is known analytically \cite{Pytte, Boozer83, meQS} that the flow
and heat flux should obey the transformations
\begin{equation}
\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
\end{equation}
and
\begin{equation}
\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
\end{equation}
respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
It was verified that the SFINCS code obeyed these isomorphism transformations
for a range of $y$, collisionality regimes, and collision operators.

>>>>

\begin{figure}[h!]
% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
%******************************
% Previous version of figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% m20130324_01_SFINCS_ErScan_2013-03-27_03-59_ErScan_awesomer_mergeder.mat
% Code run on Matt's laptop
% *********************************
\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_LHD.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.pdf}
\caption{(Color online) Comparison of trajectory models for LHD standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
of the normalized radial electric field (\ref{eq:EStar}).
\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
\label{fig:ErComparison_LHD}}
\end{figure}

\begin{figure}[h!]
% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
%******************************
% Previous version of figure generated using m20131205_05_plotSFINCSFortranErComparisonForPaper.m
% SFINCS run on hopper in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
%******************************
% Old version of figure from Dec 2 version:
% Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% sfincs_2013-12-01_15-49_ErScan_noConvergence
% Code run on edison in 
% *********************************
\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_W7X.eps}
%\includegraphics{m20131205_05_plotSFINCSFortranErComparisonForPaper_W7X.eps}
\caption{(Color online) Comparison of trajectory models for W7-X standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
of the normalized radial electric field (\ref{eq:EStar}).
\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
\label{fig:ErComparison_W7X}}
\end{figure}

@@

A useful test of a stellarator neoclassical code such as the one
described here is the quasisymmetry isomorphism,
discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
\changed{
A quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
for some periodic function $y$ and integers $M$ and $N$.
Magnetic fields with the same $y$ but different $M$ and $N$ are said to be isomorphic in that
the associated transport matrices must be related in the following manner.
Suppose the transport matrices are computed for several quasisymmetric magnetic fields with different values of $M$ and $N$,
varying the collision frequency in each calculation so
$\nuii / (\iota M - N)$ remains fixed,
and varying the radial electric field so
$(d\Phi_0/d\psi)(GM+IN)/(\iota M-N)$
remains fixed.
%Other input parameters such as $dn/d\psi$ and $dT/d\psi$ are kept fixed
%as $M$ and $N$ are varied.
%Under these circumstances, 
In such a scan of $M$ and $N$,
it can be shown analytically \cite{Pytte, Boozer83, meQS} that the 
transport matrix elements should vary as follows:
$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ vary
$\propto (NI+MG)^2/(\iota M-N)$;
$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ vary
$\propto (NI+MG)/(\iota M-N)$;
and $L_{33}$ varies
$\propto 1/(\iota M-N)$.
This isomorphism holds for all the trajectory models considered in this paper.
}
\begin{comment}
flow
and heat flux should obey the transformations
\begin{equation}
\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
\end{equation}
and
\begin{equation}
\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
\end{equation}
respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
\end{comment}

\changed{
As $\nuii / (\iota M - N)$ is to be held fixed in this test, while $\iota M - N$ can change sign as $M$ and $N$ are varied, 
the collision frequency to use can be negative.
While $\nuii <0$ does not make sense physically, it poses no mathematical or numerical problem.
If the signs of the collision frequency, $\xi$, and $d\Phi_0/d\psi$ are simultaneously
reversed in the kinetic equation, the part of $f_{i1}$ driven by $\left< E_{||}B\right>$ remains unchanged,
while the part driven by radial gradients changes sign. Thus, $L_{11}$, $L_{12}$, $L_{21}$, $L_{22}$, and $L_{33}$ change sign, while
$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ remain unchanged.
%As discussed previously, reversing the sign of $d\Phi_0/d\psi$ leaves the transport matrix unchanged in a stellarator-symmetric field.
Therefore, another way to express the quasisymmetry isomorphism 
(even for non-stellarator-symmetric $y$)
that preserves $\nuii > 0$ is the following:
%in a stellarator-symmetric $y$
if $M$ and $N$ are varied holding 
$\nuii / \left|\iota M - N\right|$ and $(d\Phi_0/d\psi)(GM+IN)/|\iota M-N|$ fixed,
$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ should vary
$\propto (NI+MG)^2/|\iota M-N|$;
$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ should vary
$\propto (NI+MG)/(\iota M-N)$;
and $L_{33}$ should vary
$\propto 1/|\iota M-N|$.
}

It was verified that the SFINCS code obeyed both versions of these isomorphism transformations
for various $y$, collisionality regimes, \changed{radial electric fields, trajectory models}, and collision operators.
